Local PCA results

### 2018-08-31 writeup/review-responses.tex

\reply{
    The same table includes correlations between MDS values for different window lengths 
    (Table \ref{tab:param_cors}). \revref
}

\begin{point}{p15}
 ``or choosing windows of equal length in base pairs rather than SNPs'' - once again.
\end{point}

\reply{
    Again, we address this in Table \ref{tab:param_cors}. \revref
}

\begin{point}{}
Using 2 PCs is common practice: only if this is the end of an analysis and the
PCA was done for visualization. Here you are using it for something so should
keep all the relevant PCs.
\end{point}

@@

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    This is an interesting and well written paper. It was a pleasant read. I have three main general
    comments:
\end{quote}

Thanks! We're glad you enjoyed it.

\begin{point}{Related work:} 
The authors provide an introduction of the main concepts, as well as some
intuition of what the method is doing and how, but I found comparison to previous approaches
to be somewhat missing. To some extent, this is due to the fact that the main goal of their
analysis is somewhat vaguely ``finding heterogeneity'', which leads to the applications of
detecting chromosomal inversions and evidence for background selection. It would help to
have a well defined set of hypotheses, test the method'۪s accuracy  using simulation (see next
comment), and compare to previous efforts in similar domains.
\end{point}

@@

\begin{point}{}
The use of weighted PCA to cope with unbalanced sample size could be better demonstrated.
Although the current explanation makes intuitive sense, this approach does not seem to be
used in previous work. The authors could design a simulation that supports their approach.
\end{point}

\reply{
    We felt that weighted PCA was a nice but unnecessary complication to this work,
    so have removed it entirely.
}

\begin{point}{}
It is conceivable that some subpopulations will have more missingness in some windows. That
may skew the resulting PCs by selecting different sample sizes for the different windows (as
discussed in Appendix B). This could distort the PCs, so that variation reflects underlying
variation in missingness. Would be good to discuss this potential issue and provide simulations.
\end{point}

\reply{
    We have tested the method against a large quantity of missing data
    in which the amount of missingness does not vary by population \llname{ll:test_missing}.
    However, missingness that is structured by population could absolutely cause variation in PCs.
    This could happen, for instance, with RAD data wherein allelic dropout is driven by
    population-specific polymorphism in the restriction site.
    The same problem to a lesser degree would be caused by mapping bias,
    and in fact we see windows around the breakpoints of the polymorphic inversions
    of the DPGP data having higher rates of missingness.
    We definitely would like our method to be insensitive to missing data,
    but if missingness is in fact correlated with the signal we are looking for 
    (changes in patterns of relatedness, as it is in these cases),
    correlation of MDS scores with amount of missingness is not in fact a problem.
    As we don't think this point is particularly helpful to the reader, 
    we have provided a plot of missingness against MDS score (Figure \ref{sfig:dpgp_missingness})
    and mentioned it in the text \revref,
    but have not added a lengthy discussion.
}

@@

\begin{point}{Appendix A:}
when using jackknife to estimate variance, each window is being divided in 10
``independent'' resampling units. Due to LD, these 10 blocks are likely correlated, which would
bias the estimates of variance. This is probably not a problem because both signal and noise
could be equally biased, but the authors may want to consider this potential issue. I wonder if
the correlation with recombination rate may be partially explained by this.
\end{point}

\reply{
    That's true, but we are unaware of a better option,
    and since the jackknife is only used to pick an appropriate window size,
    it does not seem crucial.  {\revref}
    Since results are the same across different window sizes,
    we don't think this can explain a correlation with recombination rate.
}

@@

\begin{point}{}
Is it possible to explain the results of Figure 6 just considering neutral variation in local
ancestry due to recent admixture? This may explain why ancestry seems to explain a fair
amount of variance in the lower plots of Fig 6. Local PCA has been previously used by others to
detect local ancestry blocks, e.g. see the PCAdmix approach by Brisbin et al. The authors
discuss the possibility that admixture is driving the differentiation, but do not test whether their
observations agree with neutrality.
\end{point}

\reply{
    This is a good point, but any \emph{neutral} variation in local ancestry
    is not expected to vary systematically along the chromosome,
    as we now discuss more explicitly \llname{ll:not_neutral}.
    It well could be that the patterns we see are due to selection acting differentially
    on introgressed segments,
    a scenario that would fall under ``linked selection''.
    We discuss this in the Discussion \revref.
    We also now cite and discuss the differences to PCAdmix \revreffull{1}{7}.
}

\begin{point}{}
``to remove the effect of artifacts such as mutation rate variation, we also rescale each
approximate covariance matrix to be of similar size (precisely, so that the underlying data
matrix has trace norm equal to one''. This potential issue is a bit unclear to me, since I would
expect that scaling the volume of local trees would not result in changed distances in PC
space. Perhaps the authors could show via simulations that this creates a problem, and that the
normalization addresses it.
\end{point}

\reply{
    We agree that demonstrating this issue would be nice in principle,
    but think that for this audience would be getting too far down in the weeds.
    (Besides, we aim to provide a good method, not to prove that the method
    is better than all other possible methods.)
    For the reviewer's sake, here is some more about our rationale.
    Suppose that the covariance matrix is the same along the whole genome,
    except that it is multiplied by an overall constant that varies.
    Since PCs are scaled to have norm 1, this overall scaling will indeed not affect the PCs,
    but we don't compare the PCs directly -- we compare the low-dimensional approximations
    to the covariance matrix, and the scaling we propose would remove the effect
    of this overall constant.
    It is not clear that variation in mutation rate would act like an overall multiple
    of the covariance matrix.
    Since we're interested in patterns rather than magnitudes of relatedness,
    it seems a wise thing to remove.
}

@@

\begin{point}{Application:}
Is what the authors seem to be proposing not already accounted for by linear
mixed model association approaches? If not, this should be clarified. Either way, this paragraph
could be dropped.
\end{point}

\reply{
    It is not accounted for; we've added some more explanation to this section
    (which we think should stay in). \revref
}

\begin{point}{Introduction:}
 ``it is not necessarily clear what aspects of demography should be included in the
concept.'' I find it a bit weird to describe selection as an ``aspect of demography''. Although it
could be seen as such within a coalescent framework, that seems to be just a useful
representation. The authors may consider rewording.
\end{point}

\reply{
    Good point, but we think that the sense of unease is perhaps desirable here:
    certainly differential survival is an aspect of demography;
    but as the rest of the paragraph goes on to discuss,
    this quickly spills over into selection, which probably should not be included.
    We've opted to keep this. \revref
}

@@

\begin{point}{}
``There has been substantial debate over the relative impacts of different forms of selection.''
Citation needed.
\end{point}

\reply{
    We have added some relevant recent citations,
    but apologize for not being aware of the appropriate recent review to refer to.
    \revref
}

\begin{point}{}
``Results using larger numbers of PCs were nearly identical''. It would be interesting to have a
supplementary table.
\end{point}

\reply{
    We have included a table of correlations (Table \ref{tab:param_cors}),
    as well as the plots for Drosophila data using $k=5$
    (Figure \ref{fig:dmel_k5}),
    and hope that readers will take our word on this point more generally.
}

### 2018-09-15 medicago/medicago_data_setup.Rmd
wrong comma, closes #7

Now let's see how it goes to do this for an entire chromosome.
On chromosome 1 there are about 5 million SNPs, so we'll do it in windows of 5,000.
First, windows by SNPs.
```{r lpca_chr1_snp, cache=TRUE}
sites <- vcf_positions(bcf.file)
do.indivs <- sample( samples$ID, 50 )
win.fn.snp <- vcf_windower(bcf.file, size=5e3, type="snp", sites=sites, samples=do.indivs) 
system.time( snp.pca <- eigen_windows(win.fn.snp,k=2) )
```
It takes about 2 minutes to do chromosome 1, for ten individuals, on phoebe.
Next, we compute the distance between windows with `pc_dist()` and do MDS visualization on the result.
This is just with ten samples, so we don't expect a real result, but let's check it all works:
```{r mds_chr1_snp, cache=TRUE, depends="lpca_chr1_snp"}
system.time( chr1.pcdist <- pc_dist( snp.pca ) )
# there may be windows with missing data
na.inds <- is.na( snp.pca[1,] )
chr1.mds <- cmdscale( chr1.pcdist[!na.inds,!na.inds], eig=TRUE, k=4 )
mds.coords <- chr1.mds$points[ ifelse( na.inds, NA, cumsum(!na.inds) ), ]
colnames(mds.coords) <- paste("MDS coordinate", 1:ncol(mds.coords))
```

>>>>

Now let's see how it goes to do this for an entire chromosome.
On chromosome 1 there are about 5 million SNPs, so we'll do it in windows of 5,000.
First, windows by SNPs.
```{r lpca_chr1_snp, cache=TRUE}
sites <- vcf_positions(bcf.file)
do.indivs <- sample( samples$ID, 50 )
win.fn.snp <- vcf_windower(bcf.file, size=5e3, type="snp", sites=sites, samples=do.indivs) 
system.time( snp.pca <- eigen_windows(win.fn.snp,k=2) )
```
It takes about 2 minutes to do chromosome 1, for ten individuals, on phoebe.
Next, we compute the distance between windows with `pc_dist()` and do MDS visualization on the result.
This is just with ten samples, so we don't expect a real result, but let's check it all works:
```{r mds_chr1_snp, cache=TRUE, depends="lpca_chr1_snp"}
system.time( chr1.pcdist <- pc_dist( snp.pca ) )
# there may be windows with missing data
na.inds <- is.na( snp.pca[,1] )
chr1.mds <- cmdscale( chr1.pcdist[!na.inds,!na.inds], eig=TRUE, k=4 )
mds.coords <- chr1.mds$points[ ifelse( na.inds, NA, cumsum(!na.inds) ), ]
colnames(mds.coords) <- paste("MDS coordinate", 1:ncol(mds.coords))
```
